---
title: "Stepwise Regression Pt 2"
author: "Leah Wood"
date: "2025-07-09"
output: pdf_document
---
---
title: "Stepwise Regression"
author: "Leah Wood"
date: "2025-07-02"
output: html_document
---

# Step 0
### Call in the data

```{r}
library(dplyr)
```

```{r}
#Load in the Data
codebook<-read.csv("codebook.csv", header=T)
covariates<-read.csv("covariates.csv", header = T)
exposome<-read.csv("exposome.csv", header = T)
phenotype<-read.csv("phenotype.csv", header=T)
covariates<-covariates[,-1]
phenotype<-phenotype[,-1]
proteome<-read.csv("proteome_clean.csv")
data_merge<-read.csv("data_merge.csv")
#Data merge has PFAS (mother and child), covariates, GenTot, behavior classification, and an ID
data_merge<-data_merge[,-1]
```

```{r}
#Ensure things are factors
data_merge$h_cohort = as.factor(data_merge$h_cohort)
data_merge$e3_sex_None = as.factor(data_merge$e3_sex_None)
data_merge$e3_yearbir_None = as.factor(data_merge$e3_yearbir_None)
data_merge$h_edumc_None = as.factor(data_merge$h_edumc_None)
data_merge$h_native_None = as.factor(data_merge$h_native_None)
data_merge$h_parity_None = as.factor(data_merge$h_parity_None)
```

```{r}
#Fix the proteome data
proteome<-rename(proteome, ID = X)
proteome<-proteome|>arrange(ID)
#Data merge proteome has the PFAS (mother and child), covariates, GenTot, behavior classification, and all proteins, and an ID
data_merge_proteome <- merge(proteome, data_merge, by.x = "ID", by.y = "ID")
```

# Step 1
### PFAS in the Mother to PFAS in the Child

```{r}
test1<-lm(hs_pfhxs_c_Log2~hs_pfhxs_m_Log2, data=data_merge_proteome)
summary(test1)$coefficients
```


```{r}
#This model shows how PFAS in the mother impact PFAS in the child while accounting for the mother's confounding factors.

#We would be interested to look at how PFAS interact between mother and child WITHOUT the covariates as well.

results_PFAS_mc <- data.frame(
  PFAS_ch = character(),
  PFAS_m = character(),
  Coefficient = numeric(),
  P_value = numeric(),
  stringsAsFactors = FALSE
)

for (i in 1:5) {
    test <- lm(data_merge_proteome[[2*i+36]]~data_merge_proteome[[2*i+37]]+h_mbmi_None+hs_wgtgain_None+e3_gac_None+h_age_None+h_edumc_None+h_native_None+h_parity_None, data=data_merge_proteome)
    
        results_PFAS_mc <- rbind(results_PFAS_mc, data.frame(
        PFAS_ch = names(data_merge_proteome)[2*i+36],
        PFAS_m = names(data_merge_proteome)[2*i+37],
        Coefficient = summary(test)$coefficients[2,1],
        P_value = summary(test)$coefficients[2,4]))
}


print(results_PFAS_mc)

```

The results above show that PFAS exposure in the mother significantly impacts PFAS exposure in the child after controlling for confounding factors. 

#### Figure of Step 1

```{r}
library(ggplot2)

plots <- list()  # initialize an empty list

for (i in 1:5) {
  p <- ggplot(data_merge_proteome, 
              aes(x = data_merge_proteome[[2*i + 37]], 
                  y = data_merge_proteome[[2*i + 36]])) +
    geom_point() +
    labs(x = names(data_merge_proteome)[2*i + 37],
         y = names(data_merge_proteome)[2*i + 36],
         title = paste("Plot", i))
  
  plots[[i]] <- p  # store each plot in the list
}

# Display all 5 plots (one by one)
for (i in 1:5) {
  print(plots[[i]])
}


```

# Step 2
### PFAS to Behavior 

First we need to sum the PFAS in the children to show a accumulated effect. 

```{r}
data_merge_proteome<-data_merge_proteome|>mutate(sum_c_pfas=hs_pfhxs_c_Log2+hs_pfna_c_Log2+hs_pfoa_c_Log2+hs_pfos_c_Log2+hs_pfunda_c_Log2)
```

```{r}
#We are going to add in covariates about the children to adjust for confounding factors. 
pfas_sum_behavior<-lm(hs_Gen_Tot~sum_c_pfas+h_cohort+e3_sex_None+hs_child_age_None+hs_c_weight_None+hs_c_height_None+h_age_None+h_edumc_None+h_native_None+h_parity_None, data=data_merge_proteome)
summary(pfas_sum_behavior)
#If there's time let's go back and look at other factors like drinking (e3_alcpreg_yn_None)
```

This shows that the sum of PFAs exposure in children is significantly associated with the CBCL score even after adjusting for confounding factors. 

```{r}
#We can make a heat map for this model 
library(reshape2)
library(ggplot2)
st2_cor<-cor(data_merge_proteome[,c("hs_Gen_Tot","sum_c_pfas","hs_child_age_None","hs_c_weight_None","hs_c_height_None","h_age_None")])
melt_st2<-melt(st2_cor)
ggplot(data=melt_st2, aes(x=Var1, y=Var2, fill=value))+
  geom_tile()+ 
  scale_fill_gradient2(low="darkblue", high="darkred", midpoint = 0.5, mid="white")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.40, hjust = 1))

#We need to fix these colors
```

# Step 3
### Modeling the Impact of the Sum of PFAS on Proteins

```{r}
#This is Leah's looping method
pfas_proteins<-list()

results_PFAS_proteins <- data.frame(
  Protein = character(),
  Coefficient = numeric(),
  P_value = numeric(),
  stringsAsFactors = FALSE
)

y_proteins<-names(data_merge_proteome)[2:37]
for (i in seq_along(y_proteins)){
  y<-y_proteins[[i]]
  formula<-paste(y, "~", "sum_c_pfas")
  model <- lm(as.formula(formula), data = data_merge_proteome)
  pfas_proteins[[i]]<-model
  
  results_PFAS_proteins <- rbind(results_PFAS_proteins, data.frame(
        Protein = y,
        Coefficient = summary(pfas_proteins[[i]])$coefficients[2,1],
        P_value = summary(pfas_proteins[[i]])$coefficients[2,4]))
}

signfiicant_proteins<-results_PFAS_proteins|>filter(P_value<0.05)
print(signfiicant_proteins)
```

The sum of PFAS exposure in children is significantly associated with 13 protein levels (without checking for confounders). 

```{r}
#We are going to rerun this using John's looping method
significant_proteins_2 <- data.frame(
  Protein = character(),
  Coefficient = numeric(),
  P_value = numeric(),
  stringsAsFactors = FALSE
)

for (i in 2:37) {
    test <- lm(data_merge_proteome[[i]]~ data_merge_proteome[[63]])
    

      significant_proteins_2 <- rbind(significant_proteins_2, data.frame(
        Protein = names(data_merge_proteome)[i],
        Coefficient = summary(test)$coefficients[2,1],
        P_value = summary(test)$coefficients[2,4]
      ))
  
}

significant_proteins_2|>filter(P_value<0.05)


```

We found 13 proteins significantly associated with the PFAs sum (all of the coefficients are negative).

```{r}
#Potentially add false discovery rate analysis during this step
```

# Step 4
### Significant Proteins to Behavior

```{r}
protein_names<-signfiicant_proteins$Protein
indices <- match(protein_names, names(data_merge_proteome))
print(indices)
significant_proteins_data<-data_merge_proteome[,c(3, 4, 7, 8, 9, 11, 13, 14, 15, 16, 17, 18, 19, 20, 23, 26, 27, 28, 29, 32, 33, 35, 36,48)]
```

```{r}
#Here we are doing 13 regressions (one for each significant protein on behavior)
significant_proteins_and_CBCL <- data.frame(
  Protein = character(),
  Coefficient = numeric(),
  P_value = numeric(),
  stringsAsFactors = FALSE
)

for (i in 1:23) {
    test <- lm(significant_proteins_data$hs_Gen_Tot ~ significant_proteins_data[[i]])

      significant_proteins_and_CBCL <- rbind(significant_proteins_and_CBCL, data.frame(
        Protein = names(significant_proteins_data)[i],
        Coefficient = summary(test)$coefficients[2,1],
        P_value = summary(test)$coefficients[2,4]
      ))
}

print(significant_proteins_and_CBCL)
SPCBLC<-significant_proteins_and_CBCL|>filter(P_value<0.05)
print(SPCBLC)
```

We found 7 proteins (that are impacted by PFAS) to be significantly associated with behavior.

```{r}
#Now we are going to adjust for cofounders to see if these proteins are still significant with behavior. 
protein_names2<-SPCBLC$Protein
indices2 <- match(protein_names2, names(data_merge_proteome))
print(indices2)
proteins_interest_data<-data_merge_proteome[,c(3,4,7,8,11,14,17,50:62,48)]
```

```{r}
#Here's the loop

protein_CBCL_test<-list()
significant_proteins_and_CBCL_w_cov <- data.frame(
  Protein = character(),
  Coefficient = numeric(),
  P_value = numeric(),
  stringsAsFactors = FALSE
)

for (i in 1:7) {
    test <- lm(proteins_interest_data$hs_Gen_Tot ~ proteins_interest_data[[i]]+h_cohort+e3_sex_None+hs_child_age_None+hs_c_weight_None+hs_c_height_None+h_age_None+h_edumc_None+h_native_None+h_parity_None, data=proteins_interest_data)
  protein_CBCL_test[[i]]<-test
      significant_proteins_and_CBCL_w_cov <- rbind(significant_proteins_and_CBCL_w_cov, data.frame(
        Protein = names(proteins_interest_data)[i],
        Coefficient = summary(test)$coefficients[2,1],
        P_value = summary(test)$coefficients[2,4]
      ))
}

print(significant_proteins_and_CBCL_w_cov)


```

Once adjusting for covariates, only IL1beta and IL6 are significantly associated with behavior. 

# Visualization for Poster

### Step 1 Visualization

```{r}
data_merge_proteome<-data_merge_proteome|>mutate(sum_m_pfas=log2((2^(hs_pfhxs_m_Log2)+2^(hs_pfna_m_Log2)+2^(hs_pfoa_m_Log2)+2^(hs_pfos_m_Log2)+2^(hs_pfunda_m_Log2))))
```

```{r}
plot(sum_c_pfas~sum_m_pfas, data=data_merge_proteome, main="PFAS Exposure in Mothers VS PFAs Exposure in Child", xlab="Mother Exposure", ylab="Child Exposure")
abline(lm(sum_c_pfas~sum_m_pfas, data=data_merge_proteome), col="red")
```

```{r}
#png(filename="stepone.png")
ggplot(data=data_merge_proteome, aes(x=sum_m_pfas, y=sum_c_pfas))+
  geom_point()+
  geom_smooth(method="lm", color="red")+
  labs(title="PFAS Exposure in Mothers VS PFAs Exposure in Child", x="Mother Exposure", y="Child Exposure")
#dev.off()
```

### Step Two Visualization 

```{r}
confint(pfas_sum_behavior)
```

## Step Three Visualization

```{r}
confint_PFAS <- lapply(signfiicant_proteins$Protein, function(prot) {
  model_index <- which(y_proteins == prot)
  ci <- confint(pfas_proteins[[model_index]])[2, ]  # CI for sum_c_pfas
  return(data.frame(
    Protein = prot,
    CI_lower = ci[1],
    CI_upper = ci[2]
  ))
})
confint_PFAS
```

### Step Four Visuals 

```{r, fig.height=10, fig.width=10}
library(gridExtra)
#png(filename = "stepfour.png")
p1<-ggplot(data=significant_proteins_data, aes(y=hs_Gen_Tot, x=IL1beta))+
  geom_point()+
  labs(title="IL1beta vs CBCL Score", x="IL1beta levels", y="CBCL Score")+
  geom_smooth(col="red")
p2<-ggplot(data=significant_proteins_data, aes(y=hs_Gen_Tot, x=IL6))+
  geom_point()+
  labs(title="IL16 vs CBCL Score", x="IL6 levels", y="CBCL Score")+
  geom_smooth(col="red")
grid.arrange(p1,p2, ncol=1)
#dev.off()
```

### Experimenting with Log 2

```{r}
#This is what we have
#data_merge_proteome<-data_merge_proteome|>mutate(sum_c_pfas=log2((2^(hs_pfhxs_c_Log2)+2^(hs_pfna_c_Log2)+2^(hs_pfoa_c_Log2)+2^(hs_pfos_c_Log2)+2^(hs_pfunda_c_Log2))))

data_merge_proteome<-data_merge_proteome|>mutate(test_logs=hs_pfhxs_c_Log2+hs_pfna_c_Log2+hs_pfoa_c_Log2+hs_pfos_c_Log2+hs_pfunda_c_Log2)
```

